Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2021-12-27 paper/paper.tex
new data n=100 configurations

%more concretely, an maximal independent set of a graph $G$ corresponds to a maximum clique in the complement graph of $G$,
%and the complement vertex set of an MIS corresponds to a minimum vertex cover of the graph $G$.
%In this paper, the term ``maximal'' has a different meaning from ``maximum''  which we discussed above; a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.
In this paper, we do not limit our discussion to finding one of the maximum indepedent sets and its size.
There are many other interesting problems pertaining to independent sets, such as the number of independent sets of a given size
and enumerating independent sets at a give size or what we call computing \textit{properties}.
Some of these problems are of great interest in physics applications such as the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018} \red{cite experiment when it's ready};
they can, for example, be used to understand phase transitions, to analyse the overlap gap property~\cite{Gamarnik2013, Gamarnik2019} and to identify harder graphs in an ensemble of graphs \red{cite experiment}.
Finding the number of independent sets of a given size is equivalent to computing the coefficients of independence polynomial,
which is a useful graph characteristics that can related to Euler characteristics of the independence complex~\cite{Bousquet2008,Levit2009}, partition functions~\cite{Lee1952,Yang1952} et al.
Computing independence polynomials belongs to the complexity class \#P-hard 
and only exponential algorithms~\cite{Ferrin2014} are known to compute it exactly.
There are also some interesting works on efficiently approximating the independence polynomial~\cite{Harvey2018}, but in this work, we mainly focus on computing the independence polynomial exactly.
We introduce a tensor-network based framework to compute the various properties pertaining to independent sets.
We map the independent set problem into generic tensor network contraction with specially designed tensor element algebra.
Its algorithmic complexity is similar to that of dynamic programming, i.e. scales exponentially as ${\rm tw}(G)$.
%But the tensor network approach is highly versatile so different problems can be generically solved with minimal implementation efforts without loss of efficiency.
We benchmark our algorithms by computing various independent set properties on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance on sparse graphs.
Its high performance benefits from recent advances in random tensor-network contraction for the purposes of quantum circuit simulations~\cite{Gray2021, Pan2021, Kalachev2021}.
People developed algorithms to optimize the contraction order for tensor networks with thousands of vertices,
and the optimized contraction order gives us a near optimal space complexity $2^{\sim{\rm tw}(G)}$ to contract the tensor network.
For cases when the tensors being too large to fit into a GPU, slicing technique can be used to further reduce the space consumption.
Lastly, we provide a few examples and show that this toolbox can be used to calculate the hard-square entropy constant and overlap gap property. 
Our method can also be used for finding ``maximal'' independent set properties.
Here, the term ``maximal'' has a different meaning from ``maximum''  which we discussed above;
a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.
We show how to compute properties related to maximal independent sets as well as other combinatorial optimization problems in \App{app:otherproblems}.
The implementation of our algorithm is publicly available online: ~\cite{repolink}.

@@

\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig1.pdf}
    \caption{Benchmark results for computing different properties of independent sets of a random three regular graph with different tensor element types.
    The time in these plots only includes tensor network contraction, without taking the contraction order finding and just in time compiling time into account.
    Legends are properties, data types and devices that we used in the computation, one can find the corresponding computed property in \Tbl{tbl:generictypes}.
    (a) tree width versus the number of vertices for the benchmarked graphs. 
    (b) The computing time for calculating the MIS size and for counting of the number of independent sets (ISs), the number of MISs, and the number of independent sets having size $\alpha(G)$ and $\alpha(G)-1$ (denoted as ``MISs and (MIS-1)s'').
    (c) The computing time for calculating the independence polynomials with different approaches.
    (d) The computing time for configuration enumeration, including the enumeration of all independent set configurations, a single MIS configuration, all MIS configurations, all independent set configurations having size $\alpha(G)$ and $\alpha(G)-1$,  with or without bounding the enumeration space.
    \red{suggestions for the figure: 1. the legend seems to overlap the lines, especially the large subfigure. Not sure what's a better way to include the legend there. 2. I see you guys don't typically use markers. We typically use markers to distinguish the different lines, as well as using the color. Especially, for some journals, there is a physical non-color copy, so some requires the figures to be understandable without colors.}
    \blue{If we use markers, the data poitns are too dense. I prefer not adding markers, not much people read papers with kindle or on paper.}
    \blue{I used a smaller fontsize, there are still some overlaps. Does it look better (or worse) now?}
    }
    \label{fig:benchmark}
\end{figure}

@@

With the ability to enumerate configurations, we can analyse the landscape of the target problem, like the overlap gap property~\cite{Gamarnik2013, Gamarnik2019}.
We compute all MIS configurations for two random 3 regular graph instances of size 200, and show the Hamming distance statistics in \Fig{fig:hamming}.
Some figure has single peak structure while some other has multiple peaks. Multiple peak structure indicates disconnected clusters in the configuration space.
\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig3.pdf}
    \caption{The statistics of Hamming distances between two MIS configurations for two specific three regular graph instance of size $200$.}
    \label{fig:hamming}
\end{figure}

>>>>

%more concretely, an maximal independent set of a graph $G$ corresponds to a maximum clique in the complement graph of $G$,
%and the complement vertex set of an MIS corresponds to a minimum vertex cover of the graph $G$.
%In this paper, the term ``maximal'' has a different meaning from ``maximum''  which we discussed above; a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.
In this paper, we do not limit our discussion to finding one of the maximum indepedent sets and its size.
There are many other interesting problems pertaining to independent sets, such as the number of independent sets of a given size
and enumerating independent sets at a give size or what we call computing \textit{properties}.
Some of these problems are of great interest in physics applications such as the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018} \red{cite experiment when it's ready};
they can, for example, be used to understand phase transitions, to analyse the overlap gap property~\cite{Gamarnik2013, Gamarnik2019} and to identify harder graphs in an ensemble of graphs \red{cite experiment}.
Finding the number of independent sets of a given size is equivalent to computing the coefficients of independence polynomial,
which is a useful graph characteristics that can related to Euler characteristics of the independence complex~\cite{Bousquet2008,Levit2009}, partition functions~\cite{Lee1952,Yang1952} et al.
Computing independence polynomials belongs to the complexity class \#P-hard 
and only exponential algorithms~\cite{Ferrin2014} are known to compute it exactly.
There are also some interesting works on efficiently approximating the independence polynomial~\cite{Harvey2018}, but in this work, we mainly focus on computing the independence polynomial exactly.
We introduce a tensor-network based framework to compute the various properties pertaining to independent sets.
We map the independent set problem into generic tensor network contraction with specially designed tensor element algebra.
Its algorithmic complexity is similar to that of dynamic programming, i.e. scales exponentially as ${\rm tw}(G)$.
%But the tensor network approach is highly versatile so different problems can be generically solved with minimal implementation efforts without loss of efficiency.
We benchmark our algorithms by computing various independent set properties on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance on sparse graphs.
Its high performance benefits from recent advances in random tensor-network contraction for the purposes of quantum circuit simulations~\cite{Gray2021, Pan2021, Kalachev2021}.
People developed algorithms to optimize the contraction order for tensor networks with thousands of vertices,
and the optimized contraction order gives us a near optimal space complexity $2^{\sim{\rm tw}(G)}$ to contract the tensor network.
For cases when the tensors being too large to fit into a GPU, slicing technique can be used to further reduce the space consumption.
Lastly, we provide a few examples and show that this toolbox can be used to calculate the hard-square entropy constant and overlap gap property. 
Our method can also be used for finding ``maximal'' independent set properties.
Here, the term ``maximal'' has a different meaning from ``maximum''  which we discussed above;
a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.
We show how to compute properties related to maximal independent sets as well as other combinatorial optimization problems in \App{app:otherproblems}.

@@

\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig1.pdf}
    \caption{Benchmark results for computing different properties of independent sets of a random three regular graph with different tensor element types.
    The time in these plots only includes tensor network contraction, without taking the contraction order finding and just in time compiling time into account.
    Legends are properties, algebra and devices that we used in the computation, one can find the corresponding computed property in \Tbl{tbl:generictypes}.
    (a) tree width versus the number of vertices for the benchmarked graphs. 
    (b) The computing time for calculating the MIS size and for counting of the number of independent sets (ISs), the number of MISs, and the number of independent sets having size $\alpha(G)$ and $\alpha(G)-1$ (denoted as ``MISs and (MIS-1)s'').
    (c) The computing time for calculating the independence polynomials with different approaches.
    (d) The computing time for configuration enumeration, including the enumeration of all independent set configurations, a single MIS configuration, all MIS configurations, all independent set configurations having size $\alpha(G)$ and $\alpha(G)-1$,  with or without bounding the enumeration space.
    \red{suggestions for the figure: 1. the legend seems to overlap the lines, especially the large subfigure. Not sure what's a better way to include the legend there. 2. I see you guys don't typically use markers. We typically use markers to distinguish the different lines, as well as using the color. Especially, for some journals, there is a physical non-color copy, so some requires the figures to be understandable without colors.}
    \blue{If we use markers, the data poitns are too dense. I prefer not adding markers, not much people read papers with kindle or on paper.}
    \blue{I used a smaller fontsize, there are still some overlaps. Does it look better (or worse) now?}
    }
    \label{fig:benchmark}
\end{figure}

@@

With the ability to enumerate configurations, we can analyse the landscape of the target problem, like the overlap gap property~\cite{Gamarnik2013, Gamarnik2019}.
We compute all MIS and MIS-1 configurations for two random 3 regular graph instances of size 100, and show the Hamming distance statistics in \Fig{fig:hamming}.
The multiple peak structure indicates disconnected clusters in the configuration space.
\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig3.pdf}
    \caption{The statistics of Hamming distances between two MIS or MIS-1 configurations for two random three regular graph instances of size $100$.}
    \label{fig:hamming}
\end{figure}
